---
title: "COI vs. clinical covariates"
author: "Emma Berdan"
date: "`r Sys.Date()`"
output:
   html_document:
      code_folding: hide
      df_print: paged
      highlights: pygments
      number_sections: true
      self_contained: true
      theme: default
      toc: true
      toc_float:
         collapsed: true
         smooth_scroll: true
---

# Overview

- Principal Investigator: Aditi Hazra
- Experiment: Breast Cancer patient samples

```{r, cache = FALSE, message = FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(DESeq2)
library(DEGreport)
library(ggrepel)
library(pheatmap)
library(RColorBrewer)
library(biomaRt)
library(topGO)
library(KEGGREST)
library(Category)
library(ape)
library(data.table)
library(fgsea)
library(clusterProfiler)
library(sva)
library(gprofiler2)
library(janitor)
library(ggvenn)
library(ggplot2)
ggplot2::theme_set(theme_light(base_size = 14))
opts_chunk[["set"]](
    cache = FALSE,
    dev = c("png", "pdf"),
    error = TRUE,
    highlight = TRUE,
    message = FALSE,
    prompt = FALSE,
    tidy = FALSE,
    warning = FALSE,
    fig.height = 4)

sanitize_datatable = function(df, ...) {
 # remove dashes which cause wrapping
 DT::datatable(df, ..., rownames=gsub("-", "_", rownames(df)),
                   colnames=gsub("-", "_", colnames(df)))
}
```

# Load and pre-process data

We load our data. We look at how many zeros we have per gene and see that for many genes it is very high. We set a cutoff that at most 25 samples can have zero counts.


## Get annotations
```{r get gene_annotations}
 # library(biomaRt)
#

# Get gene annotations based on reference data version used for alignment/quantification from BiomaRt

# Specify the Ensembl release ## dataset used by bcbio (check archives if needed)
#
# # List current versions of datasets
#  ensembl <- useMart(biomart = "ENSEMBL_MART_ENSEMBL")
#  datasets <- listDatasets(ensembl)
# #
# # # Identify the proper archive to use for corresponding Ensembl release
#  archives <- listEnsemblArchives()
# #
# # # # Use ensembl GRCm39
#   ensembl <- useMart(biomart = "ENSEMBL_MART_ENSEMBL",
#                     dataset = "mmusculus_gene_ensembl")
# # #
# # # ## Build biomaRt query
#  filters = listFilters(ensembl)
#  attributes = listAttributes(ensembl)
# # #
#   gene_annotations <- getBM(attributes=c('ensembl_gene_id', 'entrezgene_id',  'external_gene_name', 'description'),
#         filters = 'ensembl_gene_id',
#        values = rownames(counts),
#         mart = ensembl)
# #
# # #remove duplicated ensembl ids with different entrez gene IDs.
#  pruned_annotations = gene_annotations %>%
#    arrange(ensembl_gene_id, ensembl_gene_id) %>%
#    distinct(ensembl_gene_id, .keep_all = TRUE)
# #
#  write.csv(pruned_annotations, "/Users/emb016/Documents/HBC04572_Rosen/data/biomart.csv")
#
 pruned_annotations <- read.csv("/Users/emb016/Documents/HBC04676_Sicinski/annotation/biomart.csv") %>% .[,-1]
```

## Load Data

```{r  show-metadata}
se <- readRDS("/Users/emb016/Documents/HBC04292_Hazra/full_data/bcbio-se.rds")

metadata <- colData(se) %>% 
    as_tibble(rownames = NULL)

metrics <- metadata(se)$metrics %>% 
    left_join(metadata, by = c("sample" = "sample"))

meta_sm <- metadata %>%
  as.data.frame() %>%
  column_to_rownames("sample")


raw_counts <- assays(se)[["raw"]] 



coding_genes <- read.table("/Users/emb016/Documents/HBC04292_Hazra/lnc_and_protein_hg38.txt", header=FALSE)

raw_counts2 <- subset(raw_counts, rownames(raw_counts) %in% coding_genes$V2) 



## Remove WU104 and WU228

raw_counts3 <- raw_counts2[,-c(3 , 96)] %>% as_tibble()  %>% 
    as.matrix()

rownames(raw_counts3) <- rownames(raw_counts2)


meta <- data.frame(metadata[-c(3 , 96),])
met <- data.frame(metrics[-c(3 , 96),])

full2 <- meta %>% left_join(met, by="sample")
rownames(full2) <- full2$sample

zipdata <- read.csv("/Users/emb016/Documents/HBC04292_Hazra/meta/zipcode_to_use.csv")

full3 <- full2 %>% left_join(zipdata, by=c("cc_study_id.y" = "Study_ID"))


```




## Look at correlations of COI and other continuous values


```{r, message = FALSE,warning = FALSE}
## subset to endpoints and remove unnecessary columns
corr_data <- full3 %>% drop_na(Age_DCIS_Dx.x, BMI.x, Dx_FUtime.x)

corr_data$Race.x <- factor(corr_data$Race)
corr_data$Race.x <- gsub("1", "NHW", corr_data$Race.x)
corr_data$Race.x <- gsub("2", "Black", corr_data$Race.x)
corr_data$Race.x <- relevel(factor(corr_data$Race.x), ref = "NHW")


## Correlate every covariate with the numeric endpoint (0 or 1)
correlations <- cor(corr_data[, c("Age_DCIS_Dx.x", "BMI.x", "Dx_FUtime.x")], corr_data$z_coi_nat)

print(correlations) 
```

## Plot all three

```{r}
ggplot(corr_data, aes(y=corr_data$BMI.x, x=corr_data$z_coi_nat, color=as.factor(CaseControl.x))) + geom_point() + geom_smooth(method="lm") + xlab("COI") + ylab("BMI")
ggplot(corr_data, aes(y=corr_data$Age_DCIS_Dx.x, x=corr_data$z_coi_nat,color=as.factor(CaseControl.x))) + geom_point() + geom_smooth(method="lm") + xlab("COI") + ylab("Age_DCIS_dx")
ggplot(corr_data, aes(y=corr_data$Dx_FUtime.x, x=corr_data$z_coi_nat,color=as.factor(CaseControl.x))) + geom_point() + geom_smooth(method="lm") + xlab("COI") + ylab("Dx_FUtime")

```
## Look at COI v. Race in cases vs. controls and race in general


```{r}
ggplot(corr_data, aes(x=as.factor(Race.x) , y=z_coi_nat, color=as.factor(CaseControl.x))) + geom_boxplot() + geom_point(alpha=0.5, position = position_jitterdodge()) + ylab("COI") + xlab("Race") + scale_color_manual(values=c("#bcbddc","#756bb1")) + theme(legend.position = "None")

ggplot(corr_data, aes(x=as.factor(Race.x) , y=z_coi_nat, color=as.factor(Race.x))) + geom_boxplot() + geom_point(alpha=0.5) + ylab("COI") + xlab("Race") + scale_color_manual(values=c("#bcbddc","#756bb1")) + theme(legend.position = "None")

```





## Plot categorical variables vs. COI as boxplots

only for cases

```{r}
cases <- subset(full3, full3$CaseControl.x==1)

ggplot(cases, aes(x=as.factor(Endocrine_Therapy.x) , y=z_coi_nat, color=as.factor(Race.x))) + geom_boxplot() + geom_point(alpha=0.5, position = position_jitterdodge()) + ylab("COI") + xlab("Endocrine Therapy") + theme(legend.position = "None") + scale_color_manual(values=c("#bcbddc","#756bb1"))


ggplot(cases, aes(x=as.factor(Grade.x) , y=z_coi_nat, color=as.factor(Race.x))) + geom_boxplot() + geom_point(alpha=0.5, position = position_jitterdodge()) + ylab("COI") + xlab("Grade") + scale_color_manual(values=c("#bcbddc","#756bb1")) + theme(legend.position = "None")

ggplot(cases, aes(x=as.factor(Inv_Dx.x) , y=z_coi_nat, color=as.factor(Race.x))) + geom_boxplot() + geom_point(alpha=0.5, position = position_jitterdodge()) + ylab("COI") + xlab("Inv.dx") + scale_color_manual(values=c("#bcbddc","#756bb1")) + theme(legend.position = "None")

ggplot(cases, aes(x=as.factor(MarginStatus.x) , y=z_coi_nat, color=as.factor(Race.x))) + geom_boxplot() + geom_point(alpha=0.5, position = position_jitterdodge()) + ylab("COI") + xlab("Margin Status") + scale_color_manual(values=c("#bcbddc","#756bb1")) + theme(legend.position = "None")


ggplot(cases, aes(x=as.factor(BRCA1_result.x) , y=z_coi_nat, color=as.factor(Race.x))) + geom_boxplot() + geom_point(alpha=0.5, position = position_jitterdodge()) + ylab("COI") + xlab("Endocrine Therapy") + ylab("COI") + xlab("BRCA1_result") + scale_color_manual(values=c("#bcbddc","#756bb1")) + theme(legend.position = "None")


ggplot(cases, aes(x=as.factor(Radiation_Therapy.x) , y=z_coi_nat, color=as.factor(Race.x))) + geom_boxplot() + geom_point(alpha=0.5, position = position_jitterdodge()) + ylab("COI") + xlab("Radiation Therapy") + scale_color_manual(values=c("#bcbddc","#756bb1")) + theme(legend.position = "None")


ggplot(cases, aes(x=as.factor(Surgery_Type.x) , y=z_coi_nat, color=as.factor(Race.x))) + geom_boxplot() + geom_point(alpha=0.5, position = position_jitterdodge()) + ylab("COI") + xlab("Surgery Type") + scale_color_manual(values=c("#bcbddc","#756bb1")) + theme(legend.position = "None")


```

## GLM

```{r}
library(vegan)
library(MuMIn)
library(car)
library(lme4)

full4 <- full3 %>% drop_na(z_coi_nat,Dx_FUtime.x, Grade.x, Inv_Dx.x, Race.x, CaseControl.x)
m1 <- glm(z_coi_nat ~ Race.x*Dx_FUtime.x + Race.x*Grade.x + Race.x*Inv_Dx.x  + Race.x + Race.x*CaseControl.x, data = full4,  na.action = "na.fail")

dredge(m1)


cases <- subset(full4, full4$CaseControl.x==1)
best <- glm(z_coi_nat ~ Inv_Dx.x + Race.x + Race.x*Inv_Dx.x , data = cases,  na.action = "na.fail")
Anova(best)
```

